---
title: "Use Case \"Insurance SCR\" - Main effects"
author: "gg"
date: "`r Sys.Date()`"
bibliography: ["./literatur/references.bib"]
link-citations: yes
output: 
  bookdown::html_document2:
    highlight: rstudio 
    toc: true
    number_sections: yes
    toc_depth: 2
---

\newcommand{\v}[1]{\mathbb{V}{\left[#1\right]}}
\newcommand{\expect}[1]{\mathbb{E}{\left[#1\right]}}
\newcommand{\RR}{\mathbb{R}}

\newcommand{\obs}[2]{{#1}^{(#2)}}
\newcommand{\attr}[3]{{#1}_#3^{(#2)}}

# Introduction

While the use case arises out of a question about life insurance within
Solvency II no familiarity with either life insurance or Solvency II is
assumed or required. This report can be read and understood on a stand
alone basis.

It is the first of a set of five reports on this use case. The five reports are
* A consecutive set of three reports covering variance decomposition "report_scr_maineff", "report_scr_interaction" and "report_scr_NonParametric". Since they build on each other and the models complexity is gradually increased, it is probably best to read them in this sequence.
* A report focusing on tails and tail measures "report_scr_Tail"
* A report performing gradient analysis "report_gradient".

The latter two reports cover independent aspects and do not directly reference each other or the former three reports on variance analysis. The gradient analysis is done using PyTorch's auto-differentiation. Hence, it assumes basic familiarity with using Python runtime environments. But the script contains references to detailed installation instructions.

All other reports are pure R and should run out-of-the-box after installation of the required packages. All packages can be found on CRAN or a mirror. 


## Scope and goals of this report

The scope of the report is mainly educational. It does not provide a
best-practice approach to explainable AI in general or the
interpretation of this specific use case in particular. Its goals are
limited to show some of the challenges encountered and some of the
solutions available. Accordingly, the models discussed are quite basic and the approach to
model selection, calibration and validation somewhat casual. Hopefully
the reader is not disappointed by discovering that this first report covers only the most simple main-effects linear model. Even though these models are considered by
some as "inherently interpretable", they do raise very similar issues to
those encountered in more complex models. Generalisations to more
elaborate models are possible since the methods discussed are model
agnostic. This is evidenced by the later reports covering linear models with interactions and non-parametric models. Furthermore, starting with these models serves as a reminder that
explainability is neither a new requirement nor a desire restricted to
machine learning. Finally, it makes sense from an analysis perspective
to start with the most simple case and take it from there.

In line with those goals the R-programming is basic and avoids
idiosyncratic constructions of the language. Where reasonable, explicit
programming is preferred over the use of packages or libraries, even though a best-practice approach would avoid such reinventing the wheel. Nevertheless, in many cases examples using dedicated libraries are provided in addition to the do-it-yourself code examples. 


## The methods

Following the taxonomy of [Molnar] the methods discussed here are
*global* and *model-agnostic*. From a practical perspective this means:

-   Global: Each method produces one single set of statistics for the
    model and dataset as a whole.
-   Model agnostic: Those statistics are derived solely by using information obtained by
    the calibrated model's `predict` method/function.

That said, there are several occasions, where the white-box nature of
linear models is exploited to gain better insights.

The report focuses on variance as key statistic. The overarching idea
may be described as feature importance by variance attribution. Variance
attribution takes the familiar concept of $R^2$ as a share of variance
explained by a model and pushes this further by attributing this share
to the input variables. Variance as key statistic makes sense because:

-   It is a familiar concept even vis-a-vis non-technical stakeholders.
-   It is a natural measure of complexity: A random variable with near
    zero variance is close to constant. Variance can also be related to
    L2 norms quantifying the "wiggliness" of functions.
-   It is closely related to the very convenient square error loss
    function.

Variance attribution is useful to determine the impact and importance of
input variables in a quantitative and rigorous fashion. Being able to
answer the natural question "What are the most important inputs?" is a
key requirement when explaining a model.

## Use case and data

Since the actuarial ideas behind this use case and reasons for its
relevance are described in
[usecase](https://aktuar.de/en/practice-areas/data-science/use-cases/use_case2/Pages/default.aspx),
it suffices here to give a very condensed description which focuses
solely on those aspects which are relevant for variance attribution. As stated
above no familiarity of the actuarial raison d'etre of the use case is
necessary to follow the report.

The use case boils down to a classical approximation or regression
problem. The goal is to find an approximation $f$ to an unknown target
function $\check f=\expect{y \lvert X}$ where the sole information about $\check f$ is a dataset
$({y}^{(i)},{x}^{(i)})$ with inputs ${x}^{(i)}$ and responses ${y}^{(i)}$.
The residuals $\epsilon^{(i)} = y^{(i)}-\check{f}(x^{(i)})$ are corrupting noise; it is not possible to
observe the point evaluations $\check f({x}^{(i)})$ directly.

In addition, due to confidentiality, the actual meaning of the input
variables is unknown which prevents applying any a-priori knowledge to
the problem. Furthermore, no information about the distribution of the
errors is available for this dataset.

The use case itself imposes no natural definition of the norm for
approximation. But because we try to approximate a conditional expectation, squared error is the natural choice as loss function for calibration of the models' parameters. Accordingly everywhere in this and the subsequent reports squared error is used as loss function. 

There are in total three datasets or "portfolios" available but this
report only covers the first one.

The following first code block reads in data and the required libraries. 

```{r, echo=TRUE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo=TRUE, cache = TRUE, cache.comments=FALSE, comment=NA, out.width="90%", attr.source=".numberLines")

suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(kableExtra))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(waterfalls))
suppressPackageStartupMessages(library(gtools))
suppressPackageStartupMessages(library(sensitivity))

# seed for reproducibility
set.seed(4711)
datadir = "./data/"

# Number of portfolio (change this to 2 or 3 on your own risk)
ipf = 1

# data (.csv) files
fnm_xtrain = paste0(datadir,"Portfolio", ipf, "/train_input.csv")
fnm_ytrain = paste0(datadir,"Portfolio", ipf, "/train_result.csv")

xtrain = as.matrix(read.csv(fnm_xtrain)[,-1])

nxtrain = dim(xtrain)[1]
nrf = dim(xtrain)[2]
# names of input variables
nm_x = paste0("x", seq.int(1,nrf))
colnames(xtrain) = nm_x

# prepare response aka y
ytrain = as.matrix(read.csv(fnm_ytrain)[,-1])
```

The dataset `xtrain` consists of `r nxtrain` observations of `r nrf` attributes
each. There are no categorical variables hence all inputs are reals.

## Some statistics of the inputs

Unless explicitly mentioned otherwise all probabilities and statistics
always refer to the empirical measure as defined by the dataset `xtrain`
or subsets thereof.

### Marginal distribution

All margins of the inputs except the 7th input variable $x_7$ are
uniform. As examples the exhibit below shows the empirical CDFs of $x_1$
and $x_7$

```{r}
allrf = c(1, 7)

for(irf in allrf){
  plot(ecdf(xtrain[,irf]), main=paste0("Empirical CDF of input variable: ", irf))
}
```

### Correlation {#inputcorrelation}

The dependence structure of the inputs is not known, but empirical
correlations can be readily computed.

```{r}
cormat = cor(xtrain)
idxpair = which(lower.tri(matrix(0, nrf, nrf)), arr.ind = TRUE)
tmp = cbind(round(cormat[idxpair] * 100, 2), idxpair)
tmp = tmp[order(abs(tmp[,1]), decreasing = TRUE),]
colnames(tmp) = c("correlation", colnames(idxpair))
ntmp = 4
tmp[1:ntmp,]
```

The table above shows the top `r ntmp` correlations in percent sorted by
decreasing absolute value. Again variable $x_7$ is special because it is
the only one with significant correlations (strongly to the 1st and less
so to the 3rd and 2nd variable) while all other correlations are near
zero[^1].

[^1]: Note that this joint distribution of the input variables seems to
    contradict the statements found in the section "Underlying data set"
    in the use case description.


The relationship between $x_1$ and $x_7$ is further evidenced by the following scatter plot.

```{r}
print(
  ggplot(data.frame(x1 = xtrain[,1], x7=xtrain[,7]), aes(x=x1, y=x7)) + 
    geom_point(size=0.1) )
```

The domain is clearly non-rectangular, which creates the negative correlation. The regular patterns are due to the dataset being generated by a quasi Monte Carlo sequence.   

# Main effects model

## Model definition and fit

A main effects model is a linear model where only the input variables are used
as features. The linear main effects model is defined as
$$ f(x) = \beta_0 + \beta_1 x_1 + \cdots + \beta_d x_d.$$ As mentioned
above $d =$ `r nrf` is the number of attributes or input variables. The fit to determine the model's parameters is done by least squares on a subset chosen iid from the full
dataset.

## $R^2$ and variance explained

The Mean Squared Error (MSE) of a model $f$ on response data $y$ is defined as $\expect{(y - f)^2}.$ To make this slightly more intuitive and better comparable across different data sets and models this report uses normalised MSE, i.e $R^2$, which is defined as
$$ R^2 = 1 - \frac{ \expect{(y - f)^2}}{\v{y}}.$$

the symbols $\expect{}$ and $\v{}$ denote expectation and variance with respect to the empirical measure.  For models which include constant functions as features (i.e. with an intercept term) and are fit by least squares the MSE on the training set is equal to the residual variance  $\v{y - f}$, and the variance of $y$ can be split into a sum $\v{y}=\v{f} + \v{y-f}$. This decomposition is the paradigmatic example of a *variance decompostion* or *variance attribution*: The overall variance $\v{y}$ is split into the sum of two terms, where each term itself is the variance of a component related to the model. All attempts at variance decomposition follow this pattern.   

In this case $R^2$ simplifies to $$ R^2 = \frac{\v{f}}{\v{y}}.$$
This justifies the interpretation of $R^2$ as the *variance explained* by the model $f$. 

```{r, fig.cap="Variance decomposition into percent \"explained\" and \"residual\""}
# size of subset for the calibration
dessize = 4000
# select the points in the design
idxdes = sample.int(nxtrain, dessize, replace = FALSE)
x = xtrain[idxdes,]
y = ytrain[idxdes]
# fit the model
dfin = as.data.frame(cbind(y, x))
md = lm( y ~ ., dfin)  

# we just use the last model retained from above 
cf = md$coefficients[-1]
# variance of the response
var_y = var(y)
# variance of the fitted model, we do not need the intercept
var_f = var(x %*% cf)
relvar_f = var_f / var_y
relvar_res = 1 - relvar_f
df = data.frame(values = c(round(relvar_f * 100,1), round(relvar_res * 100,1)), 
                labels=c("explained (f)", "residual"))
p = waterfall( df, calc_total = TRUE, total_axis_text = "response (y)")
print(p)
```

It is difficult to assess from the residual variance alone whether the
model is a good or bad fit. This is because the residual variance has
two very different components: An unavoidable variance which is caused
by the random noise $\epsilon$ and a deterministic component which is
due to the lack of fit between $f$ and $\check f$. The unavoidable
variance contained in the `r paste0(round(relvar_res * 100,1),"%")`
above will remain even if we could use the true regression function $\check f$ for
prediction. So, if a large part of the residual variance is already due
to noise, the model can be considered quite good and there will be
little room for improvement by a better model. But disentangling these
two components would require detailed information about size and
structure of the noise which is not available[^2].

[^2]: Indeed it is well known that these errors are in general
    heteroskedastic. Furthermore, often certain shortcuts in the
    generation of the inner scenarios are employed to increase speed. Typical examples are antithetic variables or the shift algorithm for the initial yield curve. 
    These shortcuts may create complicated dependencies between errors
    at different outer scenarios. Unfortunately this is not covered in
    the use case description. While the use case provides additional data on error variance, this is not available for the base scenarios considered here.  
    
## In sample vs. out of sample fit

Since calibration of linear models is fast and requires no intervention from the user,
it is easy to perform a couple of them on different subsets to
test stability and detect possible overfitting.

```{r}
# number of iterations
niter = 10
rsq = array(Inf, dim=c(niter,3), dimnames = list(NULL, c("out", "in", "delta")) )
for(iiter in 1:niter){
  # select the points in the design
  tmp = sample.int(nxtrain, 2 * dessize, replace = FALSE)
  idxtmp = tmp[1:dessize]
  xtmp = xtrain[idxtmp,]
  ytmp = ytrain[idxtmp]
  
  # test data for out-of-sample
  idxtest = tmp[-(1:dessize)]
  xtest = xtrain[idxtest,]
  ytest = ytrain[idxtest]
  
  # fit the model
  mdtmp = lm( ytmp ~ ., as.data.frame(cbind(ytmp, xtmp)))
  # calculate Rsquareds
  rsq[iiter,"in"] = 1 - mean((ytmp - mdtmp$fitted.values)^2) / var(ytmp)
  rsq[iiter, "out"] = 1 - mean((ytest - predict(mdtmp, as.data.frame(xtest)))^2) / var(ytest)
}

rsq[,"delta"] = rsq[, "out"] - rsq[,"in"]
rsq = rbind(rsq,colMeans(rsq))
row.names(rsq) = c(rep("",niter), "mean")
kable_styling(kable(rsq*100, digits = 1, caption = "In and out of sample $R^2$ for repeated model calibration"
          ), bootstrap_options = "striped", full_width = F)
```

In conclusion, no overfitting is apparent. For this reason we do not need to distinguish between in and out of sample statistics for this model. Solely for reasons of simplicity the remainder of this first report will use only in sample statistics.

## Standardised regression coefficients

Linear models are sometimes deemed inherently interpretable because the
model structure is simple and easily accessible. It is natural to use
this knowledge to explain the effects of the features on the outcome.
The main effects model is the sum of $\beta_i x_i$ so a natural
candidate for the effect size of theses terms is their normalised
variance, i.e. the [standardised regression
coefficients](https://en.wikipedia.org/wiki/Standardized_coefficient)
defined as
$$ \mathit{src_i}^2 = 
  \frac{\mathbb{V}{\left[\beta_i x_i\right]}}{\mathbb{V}{\left[y\right]}}.$$
The following chart shows the decomposition for the `r nrf` variables $X_1,\ldots,X_d$  involved.

```{r , fig.cap="\"Decomposition\" using squared standardised regression coefficients "}
# empirical variance of the inputs
var_x = apply(x, 2, var)
# standard(ised) regression coefficients
stdregcf = cf^2 * var_x / var_y 

df = data.frame(values = round(c(stdregcf) * 100,1), 
                labels=c(names(stdregcf)) )
p = waterfall( df, calc_total = TRUE, total_axis_text = "???")
print(p)
```

This "decomposition" has the obvious problem that the terms do not add
up to 100%. While this could be easily fixed (or fudged?) by simply
rescaling the sum to 100%, it is important to appreciate the underlying
reason for the issue. The reason the terms don't sum up to 100% is the
correlation between the variables: The variance of a sum is equal to the
sum of variances only if variables are uncorrelated. This is not the
case here as was shown in Section \@ref(inputcorrelation). In fact
standardised regression coefficients are pure marginal statistics and
ignore completely any dependence between the input variables. Given that
$x_1$ and $x_7$ are strongly negative correlated the stand alone impact
on variance of either one is highly inflated. This is because any change
in one variable will be countered by some corresponding change in
opposite direction of the other variable thus partially cancelling their
joint impact on the variance. As a consequence this approach seems only
valid for uncorrelated input variables.

## Permutation feature importance

While standard regression coefficients have been around since the last
century, permutation feature importance was first published 2001 and is
a very popular approach in machine learning. See [Section
8.5.1](https://christophm.github.io/interpretable-ml-book/feature-importance.html)
of [Molnar] or [Section
4.2](https://scikit-learn.org/stable/modules/permutation_importance.html)
of [SciKit-learn].

Although permutation feature importance is readily available in many machine learning packages it is simple enough to be reprogrammed from scratch below. The main idea is to
measure the average increase in error when the model is evaluated on
data points where the value of a feature is permuted. Since it is not
feasible to calculate all permutations the average is approximated by
Monte-Carlo. A careful analysis of the code below or the description of
the algorithm reveals that the method is indeed model-agnostic, i.e. it
just relies on a model's `predict` method. Solely for reasons of speed
the code below does not use the `predict` method of the respective `md`
object but does the evaluation explicitly by performing the matrix
multiplication.

```{r, fig.cap="\"Decomposition\" using permutation feature importance"}
# unperturbed loss
pred_orig = md$coefficients[1] + x %*% cf
loss_orig = sum((y - pred_orig)^2)

# number of MC iterations
niter = 1000
# store for cumulated errors
err = rep(0, nrf)
# loop over features
for(ifeat in 1:nrf){
  # calculate the constant part and don't forget the intercept
  pred_noni = md$coefficients[1] + x[,-ifeat] %*% cf[-ifeat]
  # MC loop
  for(iiter in 1:niter){
    perm = sample.int(dessize)
    # include the permuted feature in the model's prediction 
    pred_neu = pred_noni + cf[ifeat] * x[perm,ifeat]
    # increase in loss due to permuted features
    delta_mse = sum((y - pred_neu)^2) - loss_orig
    err[ifeat] = err[ifeat] + delta_mse
  }
}

# normalise err
err = err / var_y / 2 / (dessize - 1) / niter

df = data.frame(values = round(c(err) * 100,1), 
                labels=c(nm_x) )
print(waterfall( df, calc_total = TRUE, total_axis_text = "???") )
```

Comparison with the prior chart reveals, that, indeed, for the linear
model permutation feature importance is identical to standardised
regression coefficients up to a constant factor of two. But this means that the
same caveats apply to the analysis based on permutation feature
importance as they apply to standardised regression coefficients.
Furthermore, the fact that permutation destroys the underlying
dependence structure of features should not come as a surprise to any
user and is not limited to the linear model or the squared error
loss.[^3]

[^3]: These issues often seem to be largely ignored by practitioners.
    See for example the discussion of variable importance for the
    disability use case in the original report (@Kopinsky). In the
    regression context these problems have been known and discussed for
    decades. The Machine Learning community apparently rediscovered them recently (@Hooker2021).

## New features

### Block allocation

Since most input variables of our use case are uncorrelated and the
trouble is caused only by some few variables, one way to work around the
problem may be to treat those troublesome features as a single block.
This block as a whole then gets variance allocated together with all
other uncorrelated variables. It was shown in \@ref(inputcorrelation)
that only the variables $x_i$ with $i\in \{1,2,3,7\}$ are correlated. We
introduce a new variable
$$ z = \beta_1 x_1 + \beta_2 x_2 +\beta_3 x_3 +\beta_7 x_7$$ and
calculate the standardised regression coefficients for the new variable
$z$ and the rest.

```{r, fig.cap="Attribution of variance on blocks of inputs"}
# detect variables with non-trivial correlation
eps = 1/100
cormat = cor(xtrain)
idx = lower.tri(cormat)
# flag those pairs with absolute correlation larger than the eps-threshold
idx_tmp = which(abs(cormat) > eps, arr.ind = TRUE)
flg = idx_tmp[,1] < idx_tmp[,2]
idx_ntcorr = idx_tmp[flg,,drop=FALSE]
# the correlated variables
idx_c = unique(c(idx_ntcorr))
# the complement, i.e. the uncorrelated ones
idx_nc = seq.int(1,nrf)[-idx_c]

# define new variable from correlated variables
zblock = x[,idx_c] %*% cf[idx_c]
src_zblock = var(zblock) / var_y
src_other = cf[idx_nc]^2 * var_x[idx_nc] / var_y
leakage = relvar_f - (src_zblock + sum(src_other) )

df = data.frame(values = round(c(src_other, src_zblock, leakage) * 100,1), 
                labels=c(names(src_other), "z", "leakage"))
print(waterfall( df, calc_total = TRUE, total_axis_text = "explained"))
```

The good news is that now the variances almost sum up to the variance of
$f$ which is `r paste0(round(relvar_f * 100,1), "%")`. The small difference, which is called leakage in the graph, is due to sampling error. When measured with the full dataset containing `r nxtrain` items, the correlations between zblock and the rest are close to zero. This is no longer true when using the smaller size of `r dessize` as was done here. This fact can be easily verified using the full training set, as is done below.

```{r, fig.cap="No leakage on the full dataset"}
# variance split on the whole dataset
varall_x = apply(xtrain, 2, var)
src_zblockall = var(xtrain[,idx_c] %*% cf[idx_c]) / var_y
src_allother = cf[idx_nc]^2 * varall_x[idx_nc] / var_y
relvar_fall = var(xtrain %*% cf) / var_y
leakage = relvar_fall - (src_zblockall + sum(src_allother) )
df = data.frame(values = round(c(src_allother, src_zblockall, leakage) * 100,1), 
                labels=c(names(src_other), "z", "leakage"))
print(waterfall( df, calc_total = TRUE, total_axis_text = "explained"))
```

Indeed, leakage is (sufficiently close to) zero when using all available data. Note that the variance of $f$ (the right black bar "explained") has changed from `r paste0(round(relvar_f * 100,1), "%")` to `r paste0(round((relvar_fall - 0.0005) * 100,1), "%")` since the dataset has changed.

But no matter which dataset we choose for calibration, the bad news remains. The block defined by $z$ is by far the most important variable. We can peek into the block to see exactly how big the problem is, but still have no way of distributing the contribution from the correlation.

```{r zblock-corr, fig.cap="Incomplete variance attribution for the z-block"}
# z-block break down

var_xinz = cf[idx_c]^2 * var_x[idx_c]
src_xinz = var_xinz / var_y
src_delta = src_zblock - sum(src_xinz)
df = data.frame(values = round(c(src_xinz, src_delta) * 100,1), 
                labels=c(names(src_xinz), "correlation") )
print(waterfall( df, calc_total = TRUE, total_axis_text = "z"))
```

### Principal components

The idea of introducing new variables can be pushed further by replacing
the correlated variables $x_1, x_2, x_3, x_7$ by uncorrelated features
$z_1, z_2, z_3, z_7$. Since correlation is a linear dependence measure
this amounts to a linear transformation of the features, which leaves
the structure of the linear model unchanged. Indeed we use as new
variables the four principal components respectively the eigenvectors of
the covariance matrix in order of their variance, $z_1$ having largest
$z_4$ smallest variance.

```{r}
# extract troublesome covariance submatrix
smcovmat = cov(x[,idx_c])
# diagonalise the matrix
svdob = svd(smcovmat)
U = svdob$u

# define the new uncorrelated features
tmp_z = x[,idx_c] %*% U

# sanity check: near zero correlations
cvtmp = cov(tmp_z)
if(any( abs(cvtmp[upper.tri(cvtmp)]) > 1e-12 )){
  stop("New features are not uncorrelated")
}

# define new feature matrix and coefficients
z = cbind(x[,idx_nc], tmp_z)
colnames(z) = c(colnames(x)[idx_nc], paste0("z", seq.int(1,length(idx_c))))
cfz = c(cf[idx_nc], cf[idx_c] %*% U)
names(cfz) = colnames(z)
# sanity check: model predictions remain the same
pred_x = x %*% cf
pred_z = z %*% cfz
if(any( abs(pred_x-pred_z) > 1e-12 )){
  stop("Predictions are not equal!")
}
```

Note the "sanity check" in lines 21-26 at the end of the code block above. It shows that the new features are
just a transformation of coordinates and do not change the model
predictions. Now that all features are uncorrelated it is straightforward to allocate
the variance:

```{r, fig.cap="Attribution with uncorrelated features", out.width="100%"}
# allocate variance on new features
src_zfeat = cfz^2 * apply(z,2,var) / var_y
leakage = relvar_f - sum(src_zfeat)

df = data.frame(values = round(c(src_zfeat, leakage) * 100,1), 
                labels=c(names(src_zfeat), "leakage") )
print(waterfall( df, calc_total = TRUE, total_axis_text = "ex-\nplained"))
```

Quite remarkably literally all variance is allocated to a single
principal component: $z_4$. This is all the more surprising as $z_4$ is
the principal component with the smallest variance among the four and
nicely illustrates the challenge of dimension reduction using principal
components only.

### Do new features increase interpretability?

Whether the introduction of new features helps in improving
explainability or whether this is just a confusing excuse for failing to
do better depends very much on the specific features and the domain of
their application. Unfortunately, in our Use Case we do not know
anything about the actual input variables and this question is
impossible to decide. It can only be illustrated by some complete
hypotheticals. For example assume the z-block were a group of variables
which have some underlying connection, such as all being biometric risk
factors. Then allocation on this block could readily be translated into
a statement such as "Biometric risk makes up
`r paste0(round(src_zblock * 100,1), "%")` of the total variance". Most
likely such a statement would be considered an explanation of what is
going on. If on the other hand the four variables making up the block
were, say, lapse rate, EUR/CHF FX rate and two bond indices, allocation
on this jumble of variables would not be considered an explanation but a
coincidence. An expressive new feature which has an intrinsic actuarial
interpretation independent from the sample and the model is arguably the
best and strongest way to understand the underlying reality and a most
convincing explanation. On the other hand, no matter what input
variables you have, you can always transform them into principal
components. But there will in general be no reason for those components
to have any meaning beyond being eigenvectors of an empirical covariance
matrix.

## Variance allocation according to the Shapley algorithm

A well known way to allocate contributions to each member of a set is
the marginal principle and its generalisation the Shapley algorithm. To apply this principle one needs to specify three items:

  1. The members which shall receive the allocations, 
  2. An overall amount which is to be allocated to each member
  3. A value function, which distributes the overall amount to the members. 
  
Here, the members are the input variables $X_1,\ldots, X_d$ and the overall amount 
is the variance $\v{f}$ of the full model. The value function must be
able to assign some of the total amount, i.e. variance, to every possible subset of members. For the case discussed here the value function is the variance from a linear model which employs the subset $\left\{X_1, \ldots, X_k\right\}$ from all available features.
This allows to define the increase in value due to inclusion of variable
$X_{k+1}$ as
$$ \Delta v(k+1) = \v{\mathcal{M}\left[(X_1,\ldots,X_{k+1})\right]} - \v{\mathcal{M}\left[(X_1,\ldots,X_{k})\right]}$$ 
where $\mathcal{M}$ denotes the response of the model calibrated using the first $k$ respective $k+1$ inputs.
An attribution just based on these marginal increases will depend on the
sequence of variables which does not make sense, since this sequence is
completely arbitrary. To fix this, the Shapley algorithm simply computes
increases for all possible permutations and finally averages over them. Observe, that for each permutation the sum of its marginal increases always adds up to the
variance of the full model. Accordingly the mean does this as well, so
attribution by the Shapley algorithm is additive. 

As a simple demonstration the allocation for the four variables
$x_1, x_2, x_3, x_7$ making up the $z$.block from Section \@ref(block-allocation) is programmed below.

```{r lmg-pedestrian}
# the required permutations
pm = permutations(length(idx_c), length(idx_c), idx_c)
nperm = dim(pm)[1]
nc = dim(pm)[2]

erg_ss = array(0, dim = dim(pm), dimnames = NULL)
colnames(zblock) = "zblock"
# fit models sequentially for each permutation
# step through all permutations
for(iperm in 1:nperm){
  idx = pm[iperm,]
  df = as.data.frame(cbind(zblock,x[,idx]))
  # sequential model fit, step through all variables in sequence of permutation
  for(ii in 1:nc){
    md_tmp = lm( zblock ~ ., df[,c(1,1 + (1:ii))])
    # remember variance for this group
    erg_ss[iperm, ii] = var(md_tmp$fitted.values)
  }
}

# calculate increase in variance
# prepend term with no variables, i.e. constant having zero sum of squares
erg_ss = cbind(rep(0, nperm), erg_ss)
# difference per row, i.e. increase of sum of squares due to new variable
inc_ss = t(diff(t(erg_ss)))
# relate increments to variables and aggregate all pertaining to one variable 
df_tmp = data.frame(variable=c(pm), inc_ss = c(inc_ss))
tmp = aggregate(inc_ss ~ variable, df_tmp, sum)
# normalise
val_lgm = tmp$inc_ss / nperm / c(var(zblock))

df = data.frame(values = round(c(val_lgm * c(var(zblock)) / var_y) * 100,1), 
                labels=c(nm_x[idx_c]) )
print(waterfall( df, calc_total = TRUE, total_axis_text = "z"))
```

This chart corresponds directly to the chart \@ref(fig:zblock-corr) where correlation was not distributed.

Due to being an average contribution, the amounts attributed by the Shapley algorithm may be somewhat less intuitive than directly attributed amounts. But Shapley values are always non-negative, add up to the whole amount, no matter the dependence and are uniquely defined by a set of axioms. See [Molnar] Section [9.5.3.1](https://christophm.github.io/interpretable-ml-book/shapley.html#the-shapley-value)^[What he describes as "Shapley values" is different from the definition used here. He does not distinguish between the value and the algorithm for allocation and uses a different value but his general statements mostly remain valid and apply to the situation here.] or (@Grömping) for a discussion in the context of variance attribution for linear models. But the Shapley algorithm has also been criticised and may produce counterintuitive results (see (@ShapleyProblems)).

Note also that we had to recalibrate the model for each subset (in line 15 of the prior code block). In that sense the method would not qualify as completely "model-agnostic". In practice this is not a restriction for the linear model as it can be quickly calibrated without manual intervention. 

The pedestrian approach above is woefully inefficient, which is why it was only demonstrated for four of the variables. With more efficient programming, larger numbers of variables can be handled. That said, the algorithm requires a runtime proportional to $2^d$, i.e. exponential time. For the Use Case the number of variables is $d=$ `r nrf` which is small enough to be no problem, as is demonstrated below by a computation for all variables using the package `sensitivity`.

```{r}
tt = lmg(x, y)
df = data.frame(values = round(c(tt$lmg) * 100,1), 
                labels=c(nm_x) )
print(waterfall( df, calc_total = TRUE, total_axis_text = "explained"))
```

Notice that the Shapley attribution produces the same numbers as before for the uncorrelated variables as well as for the variables in the $z$-block. The small difference between the explained value in the chart and the correct explained variance from earlier charts is due to rounding. There is no leakage when using the Shapley algorithm.

# References